---
title: "AFS 505 HW1"
author: "Aaron Appleby"
date: "2022-09-01"
output: html_document
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1.
a) __Engine1__


```{r Engine1}
# Create a vector with engine specifications in SI units (l, kW).
#combine or concatenate **c** function is used to create a vector.
#name vector with quotes " " and set equal to value
eng1 <- c("Displacement (cc)"=163.00, "Net Power Out (kW)"=3.60, "Net Torque (Nm)"=10.30, "Oil Capacity (l)"=0.58, "Fuel Tank Capacity (l)"=3.10)
eng1

# Inquire about vector type.
typeof(eng1)

# Inquire about vector attributes.
attributes(eng1)
#vector names are considered meta data
##playing with style to distinguish command notes and helpful/reminder notes

# Inquire about vector object's names
names(eng1)

# Examine a data object of vector
str(eng1)

# Inquire if vector has only a single data type (atomic vector)
is.atomic(eng1)
```
This is a 1 dimensional, double precision numeric (numeric), atomic vector.  The vector has a length of 5 named, numeric values.

1.
b) __Engine2__

```{r Engine2}
# Create a vector with engine specifications using data type (character).
eng2 <- c("Displacement (cc)"="163.00", "Net Power Out (kW)"="3.60", "Net Torque (Nm)"="10.30", "Oil Capacity (l)"="0.58", "Fuel Tank Capacity (l)"="3.10")
eng2

# Inquire about vector type.
typeof(eng2)

# Inquire about vector attributes.
attributes(eng2)
#vector names are considered meta data
##playing with style to distinguish command notes and helpful/reminder notes

# Inquire about vector object's names.
names(eng2)

# Examine a data object of vector.
str(eng2)

# Inquire if vector has only a single data type (atomic vector).
is.atomic(eng2)
#is an atomic vector even though R recognizes the names of the vector values as characters 
  ###(Must not be part of the vector???)
```
This is a 1 dimensional, ___character___, atomic vector. The vector has a length of 5 named character values.

By adding the quotes "" around the vector values they are read as character values rather than numeric values as Engine 1 from the previous question


2. __Family__

```{r Family}
# Create a character atomic vector with family members' titles and names.
fam <- c("Me"="Hip-P", "Wife"="Cam Bam", "Daughter"="Bindi", "Son"="Mumford")
fam
```

3. __Matrices__
a)

A bit of matrix review:
A **matrix** in R is a 2-dimensional data structure. 
A **matrix** holds a row (n) by collumn (m) grid of values
  [always in that order (row, collumn)]. 
As with atomic vectors, a matrix must contain values of just one data type.
(does name count toward data type??? ___not in vectors___)

'byrow' by default is FALSE, so **data is read in (input) column-wise first

```{r matrix}
# 'data' is a vector
# 'nrow' gives the number of rows
# 'ncol' gives the number of columns
# 'byrow' by default is FALSE, so data is read in column-wise
vect <- c(-3.3, 8.2, 19.7, -1.0, 10.5, 22.0, 1.3, 12.8, 24.3, 3.6, 15.1, 26.6, 5.9, 17.4, 28.9)
mat <- matrix(data = vect, nrow = 3, ncol = 5)
#colnames and rownames functions are used to address column and row name attributes
colnames(mat) <- c("AA", "BB", "CC", "DD", "EE")
mat

# Look at attributes.
attributes(mat)
# Create a line space in output (for readability).
cat("\n") 
# Examine attributes of matrix.
str(mat)
```

This is a two-dimensional, numeric matrix with column names.  The matrix contains 3 rows without names and 5 columns with names.

3. __Matrices__
b) arithmetic 

```{r matrix arithmetic}
# Add 3.3 to matrix
mat+3.3

# Divide matrix by 2.3
mat/2.3

# Original matrix
mat
```

An operation applied to the matrix is applied to every value within the matrix.

3. __Matrices__
  c) row and column manipulation

```{r matrix manipulation}
# Change the dimensions of the matrix.
#these save in the environment
dim(mat) <- c(5,3)
mat
cat("\n")
#prints a line space in output
str(mat)
```

This is a two-dimensional, numeric matrix.  The matrix contains 5 rows and 3 columns without names. The matrix contains the same vector as before and populates by column first.

3. __Matrices__
d) truncate

```{r matrix truncation}
# Truncate the numeric values in matrix.
#this does not save to the environment
# Create a new data set of the truncated matrix.
mat1 <- trunc(mat)
# View truncated matrix.
mat1
cat("\n")
str(mat1)
```

This is a two-dimensional, numeric matrix.  The matrix contains 5 rows and 3 columns without names. The matrix contains the same vector as before but truncates the numeric values.
