---
title: "AFS 505 HW2"
author: "Aaron Appleby"
date: "2022-09-03"
output: html_document
---

#HW2


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
1.
a) __crop vector__


```{r crop vector}
# Create a vector containing different crop names.
#combine or concatenate **c** function is used to create a vector.

crop <- c("apples", "wheat", "soybeans", "potatoes", "grapes", "cherries", "almonds", "hay", "hops")
crop
```

1.
b) __address vector__

Addressing allows for concise data reshaping, extraction and assignment in R.

```{r address}
## assignment - indexing w/ Brackets []
###Addressing locations with brackets calls vector value at vector position indicated within brackets.
###Addressing locations with brackets can also be used to make assignments, to store data in particular locations within a vector.
# Replace soybeans with oranges.
crop[3] <- "oranges"
crop
cat("\n")
#create a new vector using a subset of the previous vector
crop[c(1, 2, 4, 5, 6, 8, 9)]

cat("\n") #create line space

## Address with logical vectors & []
#You can also extract with logical vectors the same length as the vector being addressed inside [].
crop[c(T, T, F, T, T, T, F, T,  T)]
```

1.
c. __reorder vector__

```{r reorder}
# Create a new vector using a subset of the previous vector.
major <- crop[c(1, 2, 4, 5, 6, 8, 9)]
major
cat("\n")

# Reorder new vector.
major[c(1, 4, 5, 3, 7, 2, 6)]
cat("\n")

# Reorder new vector alphabetically using order()
major[order(major)]
```

2. __Matrices__
a) __positive indices__


Addressing applies to matrices and data frames in much the same way as to vectors.

```{r matrix}
# 'data' is a vector
# 'nrow' gives the number of rows
# 'ncol' gives the number of columns
# 'byrow' by default is FALSE, so data is read in column-wise
vect <- c(1, 6, 11, 16, 2, 7, 12, 17, 3, 8, 13, 18, 4, 9, 14, 19, 5, 10, 15, 20)
vect1 <- c(1:20)
mat1 <- matrix(data = vect1, nrow = 4, ncol = 5, byrow = T)
mat1
#two of the same matrices created with two different vectors and read in methods
cat("\n")
mat <- matrix(data = vect, nrow = 4, ncol = 5)
mat
cat("\n")
# As with vectors, matrices can be subsetted, addressed and assigned with square brackets. The difference is that the square brackets contain two parts **[??,??]**, addressing row and column dimensions.
mat[2:4, 3:4]
cat("\n")
# when addressing non sequential values the c() command is needed
mat[2, c(1,3,5)]
```

2. __Matrices__
b) __subset with negative indices__

As with vectors, we can also remove rows or columns with negative indices. For a given dimension, all indices must be positive or all must be negative.

```{r matrix with negative values}
#  Extract a subsets from the previous matrix using negative indices
mat[-3, ]
cat("\n")
mat[-2, c(-3,-5)]
cat("\n")
###there is more than one way to do the same thing in R
# Recreate the same matrix subsets as above
mat[c(1,2,4), 1:5]
cat("\n")
mat[c(1,3,4), c(1,2,4)]
```

3. __Matrices__
c) __replace values__

*Assign* values  to specific locations with addressing brackets [].

```{r matrix duplication and manipulation}
# Recall the last matrix
mat
cat("\n")
# Create a duplicate matrix for each step so original matrix remains unaltered
#already did this earlier
mat1
cat("\n")
# Replace 9, 14, and 19 with 88.
mat1[c(2:4), 4] <- 88
mat1
cat("\n")
# Replace 9, 14, and 19 with 23, 24, 25.
mat2 <- mat
mat2[c(2:4), 4] <- c(23:25)
mat2
cat("\n")
# Remove the value 8
mat3 <- mat
mat3[2,3] <- NA
mat3
```
