---
title: "AFS 505 HW3"
author: "Aaron Appleby"
date: "2022-09-12"
output: pdf_document
---

#HW3


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. __Simple Lists__
a) __list 1__


```{r list 1}
# Create a list containing different vectors and matricies.
##create a list with the **list()** function.
# we can construct a list by stringing together multiple objects
a = "Soils 502" 
b = c(1, 3, 4, 5) 
c = matrix(seq(1:6),3,2, byrow = T)


my_list = list(a, b, c)   
# contains copies of a, b, c
my_list
```

Addressing allows for concise data reshaping, extraction and assignment in R.

```{r extract}
# Extract matrix from list 1.
my_list[3]
cat("\n")

# Extract a and c twice
my_list[-2]
```

1.
b. __list 2__

```{r list 2}
# Named List within a list

name <- list(Last= "Williams", First= "Serena")
name

record <- c(801, 136)
#now assign names for meaning, notice these are character strings
names(record) <- c("wins", "losses")
record

grand.slam.wins <- list(Australian.Open= c(2003, 2005, 2007, 2009, 2010, 2015, 2017), French.Open= c(2002, 2013, 2015), Wimbeldon= c(2002, 2003, 2009, 2010, 2012, 2015, 2016), US.Open= c(1999, 2002, 2008, 2012, 2013, 2014))

cat("\n")
tennis_list <- list(name=name, age=c(36), record=record, grand.slam.wins=grand.slam.wins)
tennis_list

cat("\n")
str(grand.slam.wins)
```
grand.slam.wins is not structured as a data frame or matrix because it was input as a list composed of 4 numeric vectors.

2. __cbind, rbind__



Addressing applies to matrices and data frames in much the same way as to vectors.

```{r vector and matrix}
# 'data' is a vector
# 'nrow' gives the number of rows
# 'ncol' gives the number of columns
# 'byrow' by default is FALSE, so data is read in column-wise
v1 <- c(3, 5, 6, 12)
v1

cat("\n")
v2 <- c(T, F, T, T)
v2

cat("\n")
v3 <- c("A", "B", "C", "D", "E")
v3

cat("\n")
mat <- matrix(data = seq(2, 32, by = 3), nrow = 4, ncol = 5)
##default is byrow = F <- fills columns first
mat
##because there are more spots in the matrix than the length of the sequence, the sequence starts over until it runs out of space in the matrix

cat("\n")
mat1 <- matrix(data = c(1, -1, 1, 0, 0, 0, -1, 1), 2, 4, T)
mat1
##do not need to specify nrow = , ncol = , or byrow = ...can just put number of rows,columns,T
```

2. __Matrices__
a) __cbind__

create matrices using the vectors and matrices created above.

```{r cbind}
#  Extract a subsets from the previous matrices and bind them into a single matrix.
##avoid naming a column by extracting vector values

cbind(v1[1:4],v2)
cat("\n")

m3<- matrix(cbind(v3,v2), 5, 2)
m3
##as before, because the matrix is larger than the sequence of v2, the sequence starts over until it runs out of space in the matrix

cat("\n")
cbind(mat,v1[1:4])
##by just using cbind() the 6th column is named v1, the name is removed by calling the vector values using [] 
```

2. __Matrices__
c) __rbind__

create matrices using the vectors and matrices used above, but this time bind them by rows rather than by columns

```{r rbind}
rbind(mat[1:4,1:4],mat1)
cat("\n")

rbind(v1[1:4],mat1)
cat("\n")

rbind(v1,v2,v3)
```
3. 
a) __Read in Iris data__

```{r Iris}
iris <- read.csv("https://archive.ics.uci.edu/ml/machine-learning-databases/iris/iris.data", header=F)
head(iris)
#first row is not a header <- added header=F
##need to name columns
colnames(iris) <- c("sepal.length.(cm)", 
                    "sepal.width.(cm)", 
                    "petal.length.(cm)", 
                    "petal.width.(cm)", 
                    "class")
###tried to add a little style for ease of reading... Enter automatically spaces/lines text
head(iris)
```
3.
b) __summaries__

```{r summaries}
iris$class<-as.factor(iris$class)
i1 <- iris[iris$class == "Iris-versicolor",]
##the , after "Iris-versicolor" specifies the column 
summary(i1)
head(i1)
cat("\n")
i2 <- iris[iris$`petal.width.(cm)` > 1.5 | iris$`sepal.width.(cm)` > 3.0, ]
##I am not sure why the column names needed quotes ' ' when class did not... perhaps because of the () in the name... same thing was needed when spaces were in name, so I replaced them with .
i2
summary(i2)
cat("\n")

#separate by class and look at petal width
i3 <- iris[iris$class == "Iris-setosa",]
summary(i3)
i4 <- iris[iris$class == "Iris-virginica",]
summary(i4)
```

Iris petal width increases in alphabetical order (the order listed in summary()).  
As iris petal width decreases, sepal width increases.

4. __Cars__
a) __Read in cars data__

```{r cars}
car<- read.table("car.txt", header=T, stringsAsFactors=T)
##because car data is .txt not .csv and separated by something different than a comma, data can not be read.csv must be read.table
##data consists entirely of categorical data, stringsAsFactors=T turns all data into factors 
head(car)
summary(car)
```

4.
b) __remove missing values__

```{r}

#remove rows with NA positions
dim(na.omit(car))
summary(na.omit(car))
##na.omit removes rows with NA

#      dim(car[!is.na(car), ])
##for some reason this added rows (a lot of rows)
#        dim(car[is.na(car),])
##shows 10 entries with NA (that can be removed by na.omit())
```

4.
c) __remove duplicates__

```{r}
summary(car[!duplicated(car), ])
dim(car[!duplicated(car), ])
```

